---
title: "Using Ternary Plots to Visualise London's Energy Consumption"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 9999)
```

This post shows how ternary plots can be both an communicative and beautiful form of data visualisation. To create the ternary plot, we'll use the ggtern package (see http://www.ggtern.com/ for more info), as well as a function from my own JLLify package (see here: https://github.com/OwnKng/JLLify) to style the plots. 

### The Data

The data we're using comes the London Energy and Greenhouse Gas Inventory (LEGGI). This is a record of greenhouse gas emissions and energy consumption from homes, workplaces and transport within the Greater London area. It is published every year, though the most recent record we have is from 2017. Here, we will look at energy consumption only. 

```{r}
library(tidyverse)

energy <- read_csv(
  "https://data.london.gov.uk/download/leggi/2d6ee3f1-e928-48a9-8eab-01748c65ac6f/energy-consumption-borough-leggi.csv")

energy <- energy %>% janitor::clean_names()
energy <- energy %>% mutate_at(vars(k_wh, data_year), as.numeric)
```

As the data contains three major means sources of energy use, we can easily summarise what share each use represents for every London borough. 

```{r}
energy_summarised <- energy %>% 
  filter(leggi_year == 2017, sector != "Total", 
         area != "Unapportioned", 
         area != "London") %>% 
  group_by(area, sector) %>% 
  summarise(energy_consumption = sum(k_wh)) %>% 
  ungroup() %>% 
  group_by(area) %>% 
  mutate(energy_consumption = energy_consumption/sum(energy_consumption)) 

energy_summarised

```

### Visualising the Data

We can pipe the summarised data into a ggplot command to produce a rather standard filled column graph. 

```{r}

energy_summarised %>% 
  ggplot(aes(area, energy_consumption)) + 
  geom_col(aes(fill = sector), position = "fill") +
  coord_flip()

```

### Using a Ternary Plot 

Where we have three categories in our data, ternary plots are a good alternative to the standard column graphs. Using a ternary plot, we can see which share each category of energy consumption represents. It's also perhaps easier to see whether there's a general  trend in our data than with a column graph - e.g. does domestic energy consumption represent the main source of energy consumption across the London boroughs?    

The ggtern package is an extension to ggplot that enables the creation of ternary charts. As these plots have three axes, they can be difficult to read correctly. To remedy this, I've developed some light touch functions that handle the aesthetic styling of a ternary plot - applying axes colouring, shading certain zones and adding labels, etc. These functions are available in the JLLify package, which can be installed via github. 

```{r fig.height = 7, fig.width = 9, message=FALSE, warning=FALSE}
library(ggtern)

# devtools::install_github("OwnKng/JLLify")
library(JLLify)

energy_tern <-energy_summarised %>% 
  spread(sector, value = energy_consumption)

ggtern_base(x_label = "Domestic", 
            y_label = "Industrial\n &\nCommercial", 
            z_label = "Transport", 
            x_color = "#B8D8D8", 
            y_color = "#FE5F55", 
            z_color = "#EEF5DB") +
  geom_point(data = energy_tern, aes(Domestic, `Industrial and Commercial`, Transport), 
             shape = 17, size = 6, color = "white") + 
  theme_ternary_dark(x_color = "#B8D8D8", y_color = "#FE5F55", z_color = "#EEF5DB") + 
  labs(x = "%", y = "%", z = "%", 
       title = "% of energy consumption from different sources in London boroughs")

```

From this plot, we can see that in around two-thirds of London boroughs, domestic consumption accounts for the largest share of energy consumption. Industrial and commercial consumption accounts for the majority of energy usage in the other third. In one London borough - Hillingdon, where Heathrow Airport is located - transport represents about 50% of energy consumption.   

While this plot is useful for looking at the overall trend in the data, what it lacks is the ability to look at the distribution in each London borough (unlike the column graph). We can resolve this by faceting the visualisation, generating a longer, info-graphic style plot.   

```{r, fig.width = 9, fig.height = 30, echo=FALSE}

ggtern_base(x_label = "Dom.",
            x_color = "#B8D8D8",
            y_label = "Indus.",
            y_color = "#FE5F55", 
            z_label = "Trans.",
            z_color = "#EEF5DB") +
  geom_point(data = energy_tern, aes(Domestic, `Industrial and Commercial`, Transport), 
             shape = 17, size = 6, color = "white") + 
  facet_wrap(~area, ncol = 3) + 
  theme_ternary_dark(x_color = "#B8D8D8", y_color = "#FE5F55", z_color = "#EEF5DB") + 
  labs(title = "Energy Consumption by Sector", 
       subtitle = "% of Boroughs' energy consumption from domestic, industrial or transport uses",
       x = "%", y = "%", z = "%") + 
  theme(plot.title = element_text(size = 28, color = "white"), 
        plot.subtitle = element_text(size = 24, color = "white"))



```